---
title: "ML01-project"
output: html_notebook
---

# Breast Cancer data analysis

## Introduction
The phenotypes for characterisation are:

    Sample ID (code number)
    Clump thickness
    Uniformity of cell size
    Uniformity of cell shape
    Marginal adhesion
    Single epithelial cell size
    Number of bare nuclei
    Bland chromatin
    Number of normal nuclei
    Mitosis
    Classes, i.e. diagnosis

## Prepare the data
```{r}
bc_data <- read.table("breast-cancer-wisconsin.data", header = FALSE, sep = ",")
head(bc_data)
```
As we can see, the dataset has no header, so we add the collum names in order to have an easier manipulation of the data.
```{r}
colnames(bc_data) <- c("sample_code_number", "clump_thickness", "uniformity_of_cell_size", "uniformity_of_cell_shape", "marginal_adhesion", "single_epithelial_cell_size", "bare_nuclei", "bland_chromatin", "normal_nucleoli", "mitosis", "classes")
```
We also want to change the name for the response variable
```{r}
bc_data$classes[bc_data$classes=="2"] <- "benign"
bc_data$classes[bc_data$classes=="4"] <- "malignant"
head(bc_data)
```
```{r}

bc_data[bc_data=="?"]<-NA
#length(bc_data[bc_data$bare_nuclei==NA])
nrow(bc_data)-length(complete.cases(bc_data)[complete.cases(bc_data)==TRUE])
length(bc_data$bare_nuclei[is.na(bc_data$bare_nuclei)])
```
We have 16 missing data that are all in the bc_data$bare_nuclei collunm.
As the number of observation with missing data is low compared to the total number of observation, we could just ignore these 16 observations for the rest of the study and loose small amount of data. However, we could also replace the missing values by the mean of the collum. It does not change the global mean but reduces the variance.
We can also try to apply some algorith that will guess the value with the MICE library. We could also use the library Amelia but we have to make the asomption that all variables follow a multivariate law.
```{r}
bc_data<-na.omit(bc_data)
bc_data[,2:10] <- apply(bc_data[, 2:10], 2, function(x) as.numeric(as.character(x)))
#bc_data$classes <- as.factor(bc_data$classes)
nrow(bc_data)
```
Now we have a clean dataset.

## Vizualisation
```{r}
summary(bc_data)
```
```{r}
plot(bc_data[,c(-1,-11)])
```
We dont add the response variable into the scatter plot because box plots are better to display binnary variables

```{r}
attach(mtcars)
par(mfrow=c(3,3))
for (c in c(2:10)){
  boxplot(bc_data[,c]~classes,data=bc_data,xlab="classes",ylab=colnames(bc_data)[c])
}
```

## PCA
### what is PCA?
## Correlation
```{r}
library(corrplot)
```
```{r}
# calculate correlation matrix
corMatMy <- cor(bc_data[, c(-1,-11)])
corMatMy <- cor(bc_data[, -1])
corrplot(corMatMy, order = "hclust")

```

## Classification

### why do we do classification?

Choose a classifier C(X) that assigns a class label from {benign; malignant} to a future unlabeled observation X.
We want to assess the uncertainty in each classification
we want to understand the roles of the different predictors.

### Bayes Classifier

This probability is called the Bayes error rate. It is the lowest errorprobability that can be achieved by a classifier. It characterizes thedifficulty of the classification task

### K nearest neighbor

We can say that half of the information provided by the training set iscontained in the nearest neighbor (asymptotically)

### Linear Classification

### Naive Bayes

### Logistic regression


