---
title: "ML01-project"
output: html_notebook
---

# Breast Cancer data analysis

## Introduction
The phenotypes for characterisation are:

    Sample ID (code number)
    Clump thickness
    Uniformity of cell size
    Uniformity of cell shape
    Marginal adhesion
    Single epithelial cell size
    Number of bare nuclei
    Bland chromatin
    Number of normal nuclei
    Mitosis
    Classes, i.e. diagnosis

## clean the data
```{r}
bc_data <- read.table("breast-cancer-wisconsin.data", header = FALSE, sep = ",")
head(bc_data)
```
As we can see, the dataset has no header, so we add the collum names in order to have an easier manipulation of the data.
```{r}
colnames(bc_data) <- c("sample_code_number", "clump_thickness", "uniformity_of_cell_size", "uniformity_of_cell_shape", "marginal_adhesion", "single_epithelial_cell_size", "bare_nuclei", "bland_chromatin", "normal_nucleoli", "mitosis", "classes")
```
We also want to change the name for the response variable
```{r}
bc_data$classes.num=bc_data$classes
bc_data$classes.num[bc_data$classes.num=="2"] <- 0
bc_data$classes.num[bc_data$classes.num=="4"] <- 1
bc_data$classes[bc_data$classes=="2"] <- "benign"
bc_data$classes[bc_data$classes=="4"] <- "malignant"
head(bc_data)
```
```{r}
bc_data[bc_data=="?"]<-NA
#length(bc_data[bc_data$bare_nuclei==NA])
nrow(bc_data)-length(complete.cases(bc_data)[complete.cases(bc_data)==TRUE])
length(bc_data$bare_nuclei[is.na(bc_data$bare_nuclei)])
```
We have 16 missing data that are all in the bc_data$bare_nuclei collunm.
As the number of observation with missing data is low compared to the total number of observation, we could just ignore these 16 observations for the rest of the study and loose small amount of data. However, we could also replace the missing values by the mean of the collum. It does not change the global mean but reduces the variance.
We can also try to apply some algorith that will guess the value with the MICE library. We could also use the library Amelia but we have to make the asomption that all variables follow a multivariate law.
```{r}
bc_data<-na.omit(bc_data)
bc_data[,2:10] <- apply(bc_data[, 2:10], 2, function(x) as.numeric(as.character(x)))
#bc_data$classes <- as.factor(bc_data$classes)
#bc_data$classes <- as.numeric(bc_data$classes)
nrow(bc_data)
```
Now we have a clean dataset.

## Visualisation
```{r}
summary(bc_data)
```
```{r}
plot(bc_data[,2:10],col=(bc_data$classes.num+1))
```
We dont add the response variable into the scatter plot because box plots are better to display binnary variables

```{r}
attach(mtcars)
par(mfrow=c(3,3))
for (c in c(2:10)){
  boxplot(bc_data[,c]~classes,data=bc_data,xlab="classes",ylab=colnames(bc_data)[c])
}
```

## Correlation
```{r}
library(corrplot)
```
```{r}
# calculate correlation matrix
corMatMy <- cor(bc_data[, c(-1,-11)])
#corMatMy <- cor(bc_data[, -1])
corrplot(corMatMy, order = "hclust")

```

## PCA
### what is PCA?

## Classification
### why do we do classification?

Choose a classifier C(X) that assigns a class label from {benign; malignant} to a future unlabeled observation X.
We want to assess the uncertainty in each classification
we want to understand the roles of the different predictors.

### Create train subset
We divide the data into test and train dataset. We don't need to use the function scale() because the data uses already the same range ( from 1 to 10)

```{r}
n<-nrow(bc_data)
ntrain<-round(2*n/3)
ntest<-n-ntrain
train<-sample(n,ntrain)
bc_data.train<-bc_data[train,]
bc_data.test<-bc_data[-train,]
```




### Bayes Classifier

This probability is called the Bayes error rate. It is the lowest errorprobability that can be achieved by a classifier. It characterizes thedifficulty of the classification task

### K nearest neighbor
#### facts
We can say that half of the information provided by the training set is contained in the nearest neighbor (asymptotically).
Howerver, the KNN classifier breaks down for big dimention of datasets. We have 10 predictors and that is already a lot for KNN.
#### implementation
```{r}
library(FNN)
```

```{r}
classespred<-knn(bc_data.train[,2:10],bc_data.test[,2:10],factor(bc_data.train$classes),k=5)
table(bc_data.test$classes,classespred)#adjacent error matrix
```
```{r}
mean(bc_data.test$classes!=classespred)
```
#### find the optimal k
```{r}
Kmax<-15
ERR100<-c(0,Kmax)
for(k in 1:Kmax){
  classespred<-knn(bc_data.train[,2:10],bc_data.test[,2:10],factor(bc_data.train$classes),k=k)
  ERR100[k]<-mean(bc_data.test$classes!=classespred)
}
plot(1:Kmax,ERR100,type = "b")
```
```{r}
k_best <- which.min(ERR100)
k_best
```


###  Linear Discriminant Analysis (LDA)
####Theory
The parameters to estimates are πk,̂ μk,and ̂Σ

LDA uses the full likelihood based on the joint distribution of X and Y (generative model).

LDA assumes that the data are *Gaussian*. More specifically, it assumes that all classes share the same covariance matrix.

LDA finds linear decision boundaries in a K−1
dimensional subspace. As such, it is not suited if there are higher-order interactions between the independent variables.
Although LDA also has a number of parameters proportion top2, it isusually much more stable than QDA. This method is recommended when n is small.


####Implementation
```{r}
#bc_data.train.scaled<-scale(bc_data[train,c(-1,-11)])
#bc_data.test.scaled<-scale(bc_data[-train,c(-1,-11)])
```

```{r}
library(MASS)
lda.bc_data <- lda(classes~. ,data=bc_data.train[,c(-1,-12)])
#lda.bc_data
pred.bc_data.lda<-predict(lda.bc_data,newdata=bc_data.test[,c(-1,-12)])
perf <-table(bc_data.test$classes,pred.bc_data.lda$class)
print(perf)#confusion matrix
1-sum(diag(perf))/nrow(bc_data.test)#mean error rate
```
####Receiver Operating Characteristic (ROC)
In our case we have two classes therefore LDA assigns x to class 2 if xT̂Σ−1(̂μ2−̂μ1 ) >s,where the threshold *s* depends on the estimated prior probabibitie. If the prior probabilities cannot be estimated, or if the model assumption are not verified, a different threshold may give betterresult.
The Receiver Operating Characteristic (ROC) curve describes theperformance of the classifier for any value of s. By changing the value of s we can change the confusion matrix values. This allow us to prioritize the true positive rate (sensitivity) and false positive rate (1-specificity).
```{r}
library(pROC)
roc_lda<-roc(bc_data.test$classes,as.vector(pred.bc_data.lda$x))
plot(roc_lda)
```
### Quadratic Discriminant Analysis (QDA)
LDA don't assumes that all classes share the same covariance matrix. So The parameters to estimates are πk,̂ μk,and ̂Σk. Mean between classes, mean inside classe and empirical variance matrix in class k.

####Implementation
```{r}
qda.bc_data <- qda(classes~. ,data=bc_data.train[,c(-1,-12)])
pred.bc_data.qda<-predict(qda.bc_data,newdata=bc_data.test[,c(-1,-12)])
perf <-table(bc_data.test$classes,pred.bc_data.qda$class)
print(perf)#confusion matrix
1-sum(diag(perf))/nrow(bc_data.test)#mean error rate
```

### Naive Bayes
####Theory
For Naive Bayes classifers, we set the covariance matrix to diagonal matrix. This assumption means that the predictors are conditionally independent given the class variable  Y. A further simplification is achieved by assuming that the covariancematrices are diagonal and equal:Σ1=···=Σc=Σ=diag(σ21,...,σ2p).
Naive Bayes classifiers have a number of parameters proportional to p.They usually outperform other methods when p is very large.
#### Implementation
```{r}
library(naivebayes)
nb.bc_data <- naive_bayes(as.factor(classes)~. ,data=bc_data.train[,c(-1,-12)])
pred.bc_data.nb<-predict(nb.bc_data,newdata=bc_data.test[,c(-1,-12)],type="class")
perf <-table(bc_data.test$classes,pred.bc_data.nb)
print(perf)#confusion matrix
1-sum(diag(perf))/nrow(bc_data.test)#mean error rate
```


### Logistic regression
Logistic regression uses the conditional likelihood based on the conditional probabilities Pk(x) (discriminative model).
logReg models fit by maximizing the conditional likelihood, which is the likelihood function, assuming thexiare fixed. unction`(β) =logL(β)is maximized using an iterative optimizationalgorithm: the Newton-Raphson algorithm.
#### Binomial logistic regression
As we have only two classes to predict, we can use this classifier
```{r}
glm.bc_data <- glm(as.factor(classes)~. ,data=bc_data.train[,c(-1,-12)],family=binomial)
pred.bc_data.glm<-predict(glm.bc_data,newdata=bc_data.test[,c(-1,-12)],type="response")
perf <-table(bc_data.test$classes,pred.bc_data.glm>0.5)
print(perf)#confusion matrix
1-sum(diag(perf))/nrow(bc_data.test)#mean error rate
```



